---
title: "Assignment 2"
author: "David Gyaraki, Thao Le"
format: pdf
pdf:
  documentclass: article
  cite-method: biblatex
editor: visual
include-in-header:
  text: |
    \addtokomafont{disposition}{\rmfamily}
    \usepackage{amsmath}
    \newcommand{\bm}{\symbf}
    \newcommand{\T}{\text{T}}
    \newcommand{\pl}{\text{plim}}
    \usepackage{fancyvrb}
pdf-engine: xelatex
cap-location: top
toc: true
toc-title: Contents
number-sections: true
mainfont: Helvetica
setspace:
  linestretch: 1.25
fig-align: center
table-align: center
fig-pos: H
table-pos: H
execute:
  echo: true
  warning: false
  eval: true
code-line-numbers: false
colorlinks: true
code-block-bg: darkgray
df-print: default
highlight-style: arrow-dark
biblio-title: References
---

\clearpage

```{r setup}
# load packages
if(!require(pacman)){install.packages("pacman")}

p_load(devtools,tidyverse,dplyr,ggplot2,latex2exp,
       sampleSelection, quantreg)

#load data
dfData = read.csv("assignment2a_2023.csv")
attach(dfData)
```
\section{Question 1}
\subsection{(i)}
```{r quantile plot}
# Get the quantile values
quant=quantile(lntotexp, seq(0.1, 0.9, by=.4))

# Histogram of log of total medical expenditure
hist(lntotexp)

# Quantile plot of log of total medical expenditure
n = length(lntotexp)
plot((1:n - 1)/(n - 1), sort(lntotexp), type="l",
main = "Quantiles for log of total medical expenditure",
xlab = "Sample Fraction",
ylab = "Sample Quantile")
abline(h=quant, col = c("dark green","red", "blue"))
```
In the quantile plot, the median is indicated by the red line, the 10^{th} and 90^{th} quantile are indicated by the blue and green lines.

We can see from the distribution of log of total medical expenditure that there are few values from 0 to 4. Thus, the quantile plot increases quickly in this region. From 4 to 6, we see an increase frequencies of observations, thus, the quantile plot increases slower. The most rapid increase in the quantile plot is observed between 6 and 10, which makes sense because that is the region where most observations lie. After 10, there are less observations and the quantile plot increases rapidly again.

Although the quantile plot increases rapidly in both regions from 0 to 4 and 10 to 12, we observed a much steeper increase from 0 to 4, thus, we can say that the distribution of log total medical expenditure is left-skewed. This is confirmed by looking at its histogram.
\subsection{(ii)}

```{r quant_regression}
# Quantile regression
q= c(0.1,0.25,0.5,0.75,0.9)
quant_reg = rq(lntotexp ~ . , tau = q, data = dfData)
summary(quant_reg)

# OLS Regression
OLS_reg = lm(lntotexp ~ . , data = dfData)
summary(OLS_reg)
```

