---
title: "Assignment 1"
author: "David Gyaraki, Thao Le"
format: pdf
pdf:
  documentclass: article
  cite-method: biblatex
editor: visual
include-in-header:
  text: |
    \addtokomafont{disposition}{\rmfamily}
    \usepackage{amsmath}
    \newcommand{\bm}{\symbf}
    \newcommand{\T}{\text{T}}
    \newcommand{\pl}{\text{plim}}
    \usepackage{fancyvrb}
pdf-engine: xelatex
cap-location: top
toc: true
toc-title: Contents
number-sections: true
mainfont: Helvetica
setspace:
  linestretch: 1.25
fig-align: center
table-align: center
fig-pos: H
table-pos: H
execute:
  echo: true
  warning: false
  eval: true
code-line-numbers: false
colorlinks: true
code-block-bg: darkgray
df-print: default
highlight-style: arrow-dark
biblio-title: References
---

\clearpage

```{r setup}
# load packages
if(!require(pacman)){install.packages("pacman")}

p_load(devtools,tidyverse,dplyr,ggplot2,latex2exp,cowplot,tseries,sampleSelection)

#load data
dfData = read.csv("assignment1_2023.csv")
attach(dfData) 
```
\section{Question 1}
\subsection{(i)}
```{r}
lm_model = lm(logwage ~ age + agesq + schooling, data = dfData)
summary(lm_model)
```
\subsection{(ii)}

The sample selection problem here is to choose observations of the non-employed, which are those who have no income. The selection equation is then:
\begin{equation*}
    I_i = 
    \begin{cases}
    1 \text{ if logwage} > 0\\
    0 \text{ otherwise},
    \end{cases} 
\end{equation*}
and the second regression equation is:
\begin{equation*}
    Y_i^* = \bm{X_i'}\bm{\beta} + U_i.
\end{equation*}
We select a sample consisting of:
\begin{equation*}
    Y_i = 
    \begin{cases}
    Y_i^* \text{ if } I_i = 1\\
    \text{missing} \text{ if } I_i = 0,
    \end{cases} 
\end{equation*}

An OLS may fail in this context because the dependent variable (logwage) is missing for the non-employed sample, thus, it is not possible to derive an estimate of this variable for the non-employed
\subsection{(iii)}
The exclusion restriction variable is one that is included in $\bm{Z_i}$ but excluded from $\bm{X_i}$, I would choose 'married' as a suitable candidate for the sample selection model. My motivation is that married people tends to have stable income, and thus, employed.
``` {r Heckman}
# Create I variable:
dfData = mutate(dfData, vI = if_else(logwage > 0, TRUE, FALSE)) 
dfData["vI"][is.na(dfData["vI"])] <- FALSE

# Heckman model with restriction
heckman_rest = heckit( vI ~ married+age + agesq + schooling, logwage ~ age + agesq + schooling, data = dfData)
summary(heckman_rest)
```
```{r Heckman 2}
# Heckman model without restriction
heckman_unrest = heckit( vI ~ married+age + agesq + schooling, logwage ~ age + agesq + schooling + married, data = dfData)
summary(heckman_unrest)
```

- STILL NEED TO COMPARE OUTCOMES
\subsection{(iv)}
```{r ML}
# Maximum likelihood estimator, restricted
ML_rest = selection(vI ~ married+age + agesq + schooling, logwage ~ age + agesq + schooling, data = dfData)
summary(ML_rest)

# Maximum likelihood estimator, unrestricted
ML_unrest = selection(vI ~ married + age + agesq + schooling, logwage ~ age + agesq + schooling + married, data = dfData)
summary(ML_unrest)

```
- STILL NEED TO COMPARE OUTCOMES
\subsection{(v)}
get fitted values => plot histogram
```{r }
predicted_income <- fitted(ML_rest)  
hist(predicted_income)
```
The distribution is relatively normal?

\section{Question 2}
\subsection{(i)}
```{r schooling}
schooling_lm = lm(logwage ~ schooling, data = dfData)
summary(schooling_lm)
```
The summary of the model shows that the F-statistics is not significance, thus, we cannot interpret the effect of schooling on employment.

\subsection{(ii)}
```{r }
# Using distance as instrument variable
model1 = lm(schooling ~ distance)
X.hat.1 = fitted.values(model1)

model2 =lm(logwage ~ X.hat.1)
summary(model2)
```
```{r }
# Using subsidy as instrument variable
model3 = lm(schooling ~ subsidy)
X.hat.3 = fitted.values(model3)

model4 =lm(logwage ~ X.hat.3)
summary(model4)
```
```{r }
# Using subsidy and distance as instrument variable
model5 = lm(schooling ~ subsidy+distance)
X.hat.5 = fitted.values(model3)

model6 =lm(logwage ~ X.hat.5)
summary(model6)
```
\subsection{(iii)}
I would use only subsidy as the instrument variable. 