---
title: "Assignment 4"
author: "David Gyaraki, Thao Le"
format: pdf
pdf:
  documentclass: article
  cite-method: biblatex
editor: visual
include-in-header:
  text: |
    \addtokomafont{disposition}{\rmfamily}
    \usepackage{amsmath}
    \newcommand{\bm}{\symbf}
    \newcommand{\T}{\text{T}}
    \newcommand{\pl}{\text{plim}}
    \newcommand{\brefsection}[1]{Section \textcolor{blue}{\ref{#1}}}
    \newcommand{\beqref}[1]{Equation \textcolor{blue}{\eqref{#1}}}
    \newcommand{\breftable}[1]{Table \textcolor{blue}{\ref{#1}}}
    \newcommand{\breffig}[1]{Figure \textcolor{blue}{\ref{#1}}}
    \usepackage{fancyvrb}
    \usepackage{dcolumn}
    \usepackage{fvextra}
    \DefineVerbatimEnvironment{Highlighting}{Verbatim}{
      showspaces = false,
      showtabs = false,
      breaklines,
      commandchars=\\\{\}
    }
pdf-engine: xelatex
cap-location: top
toc: true
toc-title: Contents
number-sections: true
mainfont: Arial
setspace:
  linestretch: 1.25
fig-align: center
table-align: center
fig-pos: H
table-pos: H
execute:
  echo: true
  warning: false
  eval: true
code-line-numbers: false
colorlinks: true
code-block-bg: darkgray
df-print: default
highlight-style: arrow-dark
biblio-title: References
---

\clearpage

```{r setup q1, echo=TRUE, results='hide'}
# load packages
if(!require(pacman)){install.packages("pacman")}

p_load(devtools,tidyverse,dplyr,ggplot2,latex2exp,
       sampleSelection, quantreg, plm, nlme, knitr,car, ivreg,stargazer)
```

\section{Question 1}

\begin{table}[H] \centering
\begin{tabular}{l|cc|cc}
\hline
\hline
Judge  & \multicolumn{2}{c}{Jones} & \multicolumn{2}{c}{Smith} \\
Sentences       & Prison             & Other            & Prison        & Other          \\ \hline
Cases & 70\%                 & 30\%                & 40\%                & 60\%                \\
Future arrests   & 40\%                  & 60\%                 & 20\%               & 50\%             \\
\hline
\hline
\end{tabular}
\end{table}

\subsection{(i)} \label{q1_1}

We can treat the following problem as follows: $Y_i$ is the outcome 


\subsection{(ii)} \label{q1_2}




\subsection{(iii)} \label{q1_3}




\section{Question 2}


\subsection{(i)} \label{q2_1}
From what is given, we have $MDE = 0.1$, the power $p=0.7$, the proportion of students in control group is $p=0.5$. Assume the variance $\sigma^2=1$ To get the number of students the teacher should include in the experiment, we use the following formula:

\begin{gather} \label{eq_nmin}
\begin{aligned}
n &= \biggl( \frac{t_{1-\alpha/2} - t_{1-q}}{MDE} \biggl)^2 \frac{\sigma^2}{p(1-p)}\\
&= \biggl( \frac{1.960 + 0.524}{0.1} \biggl)^2 \frac{1}{0.5(1-0.5)}\\
&\approx 2468
\end{aligned}
\end{gather}

Thus, the teacher should include at least 2468 students in the experiment.
\subsection{(ii)} \label{q2_2}
This will change the proportion of students in treatment to $p=0.5 \times 20\% = 0.1$, using the formula in \beqref{eq_nmin}, the number of students required to participate in the experiment is:

\begin{gather}
\begin{aligned}
n &= \biggl( \frac{1.960 + 0.524}{0.1} \biggl)^2 \frac{1}{0.1(1-0.9)}\\
&\approx 6856
\end{aligned}
\end{gather}

Thus, the number of students required to participate in the experiment increases by $6856-2468=4388$ students.
\subsection{(iii)} \label{q2_3}

\section{Question 3}


\subsection{(i)} \label{q3_1}

```{r}
# Load data
dfData = read.csv("AngristEvans80.csv")
attach(dfData)

# Fraction of girls among the first born child
count_girl1 = table(dfData$SEXK)
fraction_girl1= count_girl1[[2]]/(count_girl1[[1]]+count_girl1[[2]])

# Fraction of girls among the second born child
count_girl2 = table(dfData$SEX2ND)
fraction_girl2= count_girl2[[2]]/(count_girl2[[1]]+count_girl2[[2]])

cat("Fraction of girls among the first born child is: ",fraction_girl1,"\n", "Fraction of girls among the second born child is: ",fraction_girl2)
```

```{r}
#Regress gender of second child on gender of first child
lm_second_first = lm(SEX2ND~SEXK, data = dfData)
summary(lm_second_first)
```

\subsection{(ii)} \label{q3_2}
```{r}
# First stage regression
lm_first_stage = lm(CHILD3 ~ SAMESEX, data= dfData)
summary(lm_first_stage)
```
Is the instrumental variable sufficiently strong? => yes

```{r}
# Regress number of children on whether the first two children have the same gender
lm_total = lm(KIDCOUNT ~ SAMESEX, data= dfData)
summary(lm_total)
```


\subsection{(iii)} \label{q3_3}
In this study, the treatment group includes those who have a third child and the control group includes those who have two children or less. The variables that affect decision for mothers to be assigned into treatment or control group is $Z= SAMESEX$, indicating whether the first two child are of the same sex or not.

The always takers are those who have a third child regardless of whether the first two children is of the same sex or not.
```{r}
df_always = dfData[dfData$CHILD3 == 1 & dfData$SAMESEX == 0,]
cat("The share of always takers is: ", nrow(df_always)/nrow(dfData))
```

The compliers are those who only have a third child if the first two kids are of the same sex.
```{r}
df_compliers1 = dfData[dfData$CHILD3 == 1 & dfData$SAMESEX == 1,]
df_compliers0 = dfData[dfData$CHILD3 == 0 & dfData$SAMESEX == 0,]
cat("The share of compliers is: ", (nrow(df_compliers1)+nrow(df_compliers0))/nrow(dfData))
```
The never takers are those who will never have the third child regardless of whether the first two chidren are of the same sex or not.
```{r}
df_never = dfData[dfData$CHILD3 == 0 & dfData$SAMESEX == 1,]
cat("The share of never takers is: ", nrow(df_never)/nrow(dfData))
```
Lastly, the defiers are those who will have a third child if the first two kids are of different sexes and will not have a third child if the first two kids are of the same sex. We cannot observe this as they are divided among the always taker and never taker's group.


\subsection{(iv)} \label{q3_4}
```{r}
iv_reg_hour <- ivreg(HOURSM ~ CHILD3 | SAMESEX, data = dfData)
iv_reg_income <- ivreg(INCOME1M ~ CHILD3 | SAMESEX, data = dfData)

stargazer(iv_reg_hour, iv_reg_income,
          type="text", report="vc*stp",
          keep.stat=c("n","adj.rsq"),
          title = "...") #remember to change title
```


\subsection{(v)} \label{q3_5}
```{r}
# Subgroup 1: Always taker
hour1=mean(df_always$HOURSM)
income1=mean(df_always$INCOME1M)
cat("The mean working hour of always takers is: ", hour1, ", the mean income of always takers is: ",income1)
```
```{r}
# Subgroup 2: never takers
hour2=mean(df_never$HOURSM)
income2=mean(df_never$INCOME1M)
cat("The mean working hour of never takers is: ", hour2, ", the mean income of never takers is: ",income2)
```
```{r}
# Subgroup 3: complier 1
hour3=mean(df_compliers1$HOURSM)
income3=mean(df_compliers1$INCOME1M)
cat("The mean working hour of complier in treatment group is: ", hour3, ", the mean income of this group is: ",income3)
```
```{r}
# Subgroup 4: complier 0
hour4=mean(df_compliers0$HOURSM)
income4=mean(df_compliers0$INCOME1M)
cat("The mean working hour of complier in control group is: ", hour4, ", the mean income of this group is: ",income4)
```

To-dos: USE these means to say something about the preference of having a third child

\subsection{(vi)} \label{q3_6}



\subsection{(vii)} \label{q3_7}
First, we stratify the sample by gender of the first child:
```{r }
df_first_girl = dfData[dfData$SEXK == 1,]
df_first_boy = dfData[dfData$SEXK == 0,]
```


(But they ask to use the first stage result?) I try to to it manually below:

```{r}
# First stage regression
lm_1st_girl = lm(CHILD3 ~ SAMESEX, data=df_first_girl)
lm_1st_boy = lm(CHILD3 ~ SAMESEX, data=df_first_boy)
stargazer(lm_1st_girl, lm_1st_boy,
          type="text", report="vc*stp",
          keep.stat=c("n","adj.rsq"),
          title = "...") #remember to change title

```
If the first child is a girl and first two children are of the same sex, one is more likely to have a third child.

Then, we perform instrumental variable regressions:
```{r}
iv_reg_girl_hour <- ivreg(HOURSM ~ CHILD3 | SAMESEX, data = df_first_girl)
iv_reg_girl_income <- ivreg(INCOME1M ~ CHILD3 | SAMESEX, data = df_first_girl)
iv_reg_boy_hour <- ivreg(HOURSM ~ CHILD3 | SAMESEX, data = df_first_boy)
iv_reg_boy_income <- ivreg(INCOME1M ~ CHILD3 | SAMESEX, data = df_first_boy)
stargazer(iv_reg_girl_hour, iv_reg_girl_income,iv_reg_boy_hour,iv_reg_boy_income,
          type="text", report="vc*stp",
          keep.stat=c("n","adj.rsq"),
          title = "...") #remember to change title

```
Here we see that if the first child is a girl, having a third child does not significantly influence the hour and income.