---
title: "Assignment 5"
author: "Group name: Foodies with hoodies"
format: pdf
pdf:
  documentclass: article
  cite-method: biblatex
editor: visual
include-in-header:
  text: |
    \addtokomafont{disposition}{\rmfamily}
    \usepackage{amsmath}
    \newcommand{\bm}{\symbf}
    \newcommand{\T}{\text{T}}
    \newcommand{\pl}{\text{plim}}
    \newcommand{\brefsection}[1]{Section \textcolor{blue}{\ref{#1}}}
    \newcommand{\beqref}[1]{Equation \textcolor{blue}{\eqref{#1}}}
    \newcommand{\breftable}[1]{Table \textcolor{blue}{\ref{#1}}}
    \newcommand{\breffig}[1]{Figure \textcolor{blue}{\ref{#1}}}
    \usepackage{fancyvrb}
    \usepackage{dcolumn}
    \usepackage{fvextra}
    \DefineVerbatimEnvironment{Highlighting}{Verbatim}{
      showspaces = false,
      showtabs = false,
      breaklines,
      commandchars=\\\{\}
    }
pdf-engine: xelatex
cap-location: top
toc: true
toc-title: Contents
number-sections: true
mainfont: Arial
setspace:
  linestretch: 1.25
fig-align: center
table-align: center
fig-pos: H
table-pos: H
execute:
  echo: true
  warning: false
  eval: true
code-line-numbers: false
colorlinks: true
code-block-bg: darkgray
df-print: default
highlight-style: arrow-dark
biblio-title: References
---

\clearpage

```{r setup q1, echo=TRUE, results='hide'}
# load packages
if(!require(pacman)){install.packages("pacman")}

p_load(devtools,tidyverse,dplyr,ggplot2,latex2exp,stargazer, fixest, modelsummary, knitr)
```

\section{Question 1}

\subsection{(i)} \label{q1_1}

Suppose the differences in outcomes between the treatment and the control group is:

```{=tex}
\begin{equation} \label{eq_1}
Y_{g1} - Y_{g0}=(\alpha_1-\alpha_0) + \delta D_g + (U_{g1} - U_{g0}),
\end{equation}
```
\noindent in which: $\delta$ is the estimated treatment effect.

The parallel trends assumption state that without the intervention of the treatment $(\delta =0)$, the difference of between the control and treatment group ($\alpha_1-\alpha_0$) remain constant over time. Since in this example, they only look at the pre-treatment period, the parallel trends assumption could be violated due to the fact that after the treatment period, the differences in outcomes of control and treatment groups are not constant over time anymore unrelated to the treatment.

When parallel trend is violated, it means that $\alpha_1-\alpha_0$ changes over time and this means it is no longer possible to estimate Equation \ref{eq_1} using OLS. If we continue estimating it using OLS, we will have a biased and inconsistent estimator.

One example for violation of this assumption could be that the there exists autocorrelation in the treatment group after getting treated. Specifically, the outcome of the next time lag is influenced by the outcome of the previous lag. Prior to the treatment, both the control and the treatment group have the same time trend because both experience no treatment. However, after receiving the treatment but not due to this, the treated group has a steeper slope in their outcomes due to autocorrelation and is no longer parallel to the control group. Suppose for example that one wishes to investigate the treatment effects on savings among the poorest income bracket, where treatment is giving them a fixed amount of monthly stipend. However, the pre-treatment period was during a crisis, where the poorest income bracket savings were stagnating. After the treatment, there was a booming economic period, where savings were autocorrelated. Then one group turns out to experience a divergent trend in savings not due to the stipend but simply because the economic situation changed the dependent variable's properties.

\subsection{(ii)} \label{q1_2}

The difference-in-difference estimator is an OLS estimator Equation \ref{eq_1}, which can be written in the form below:

```{=tex}
\begin{equation} \label{eq_2}
Y_{g1} - Y_{g0}=\beta_0 + \delta D_g + U_g.
\end{equation}
```

Now suppose that $Corr(D_g,\beta_0) \neq 0$. The main problem with applying the OLS estimator in this case is that the estimator will be biased and inconsistent due to endogeneity issues. This is because the estimator treats the time trend $\beta_0$ as a constant, however, in this case, the parallel trends assumption is not satisfied and thus $\beta_0$ is not constant. This makes the error term change over time and is correlated with the treatment variable $D_g$. This implies that the treatment was probably not assigned time-group conditional randomly and this causes the endogeneity of the treatment to the error term, which in turn causes the previously mentioned biasedness and inconsistency.

\section{Question 2} \label{q2}

```{r}
dfData = read.csv("assignment5.csv")
attach(dfData)
```

\subsection{(i)} \label{q2_1}

```{r}
# Get subgroups and mean per year
df_scarlet = dfData[dfData$treated==1,]
mean_scarlet = df_scarlet %>% group_by(year) %>% 
  summarise(mean_rate=mean(lnm_rate),
            .groups = 'drop')
df_tuber = dfData[dfData$treated==0,]
mean_tuber = df_tuber %>% group_by(year) %>% 
  summarise(mean_rate=mean(lnm_rate),
            .groups = 'drop')

# reformat into dataframe
df_grouped = data.frame(mean_tuber$year,mean_scarlet$mean_rate,mean_tuber$mean_rate)
names(df_grouped) = c("year","mean_scarlet","mean_tuber")

# Plot
ggplot() +                    
  geom_line(data=df_grouped,aes(y=mean_scarlet,x= year,colour="Scarlet"),size=1 )+
  geom_line(data=df_grouped,aes(y=mean_tuber,x= year,colour="Tuber"),size=1) +
  scale_color_manual(name = "Disease", values = c("Tuber" = "darkblue", "Scarlet" = "red")) + 
  geom_vline(xintercept = 1937) #add a vertical line indicating treatment year
```

From our graph, we can observe that while both diseases had a relatively slight downwards trend (so at first glance, we may assume parallel trends), the treatment appearance seems to affect the scarlet fever with a strong drop-off in means. This might imply that the drop in the scarlet fever against the tuberculosis is due to the treatment (the appearance of the drug is the treatment for fever but not a treatment for tuberculosis). However, we have to make a lot of assumptions for this to indeed hold true.

\subsection{(ii)} \label{q2_2}

```{r}
# get mean effects
mean_treated_1936 = as.numeric(df_scarlet[df_scarlet$year==1936,] %>% 
  summarise(mean_rate=mean(lnm_rate),
            .groups = 'drop'))
mean_treated_1937 = as.numeric(df_scarlet[df_scarlet$year==1937,] %>% 
  summarise(mean_rate=mean(lnm_rate),
            .groups = 'drop'))

mean_control_1936 = as.numeric(df_tuber[df_tuber$year==1936,] %>% 
  summarise(mean_rate=mean(lnm_rate),
            .groups = 'drop'))
mean_control_1937 = as.numeric(df_tuber[df_tuber$year==1937,] %>% 
  summarise(mean_rate=mean(lnm_rate),
            .groups = 'drop'))

Before_Treatment_1936 <- c(mean_treated_1936, mean_control_1936, (mean_treated_1936-mean_control_1936))
After_Treatment_1937 <- c(mean_treated_1937, mean_control_1937, (mean_treated_1937-mean_control_1937))
Difference <- c((mean_treated_1936 - mean_treated_1937), (mean_control_1936 - mean_control_1937), ((mean_treated_1936 - mean_treated_1937) - (mean_control_1936 - mean_control_1937)))

dfTable <- data.frame(Before_Treatment_1936, After_Treatment_1937, Difference)
rownames(dfTable) <- c("Treatment", "Control", "Difference")

kable(dfTable, caption="Treatment and time differences of treatment and control groups", digits=3, label = "tab_did") 
```

From Table 1, we can see that the difference-in-differences estimator is $0.439$. This DiD estimator is obtained by the differences in means, for treatment group pre-treatment, treatment group post-treatment, control group pre-treatment and control group post-treatment.

```{=tex}
%\begin{table}[]
%\centering
%\begin{tabular}{|l|l|l|l|}
%\hline
%           & Before treatment (1936) & After treatment (1937) & Difference \\ \hline
%Treatment  & -10.96                  & -11.43                 & 0.47       \\ \hline
%Control    & -7.61                   & -7.63                  & 0.02       \\ \hline
%Difference & -3.35                   & -3.8                   & 0.45       \\ \hline
%\end{tabular}
%\end{table}
```

\subsection{(iii)} \label{q2_3}

```{r}
# Create indicator variable
dfData$indicator <- ifelse(dfData$year >=1937, 1, 0)

# Get subdata for the year 1936 and 1937
dfSub = dfData[dfData$year==1936 | dfData$year==1937,]

# DiD model
DiD1 = feols(lnm_rate ~ indicator*treated| year + treated, data = dfSub, se="standard")
summary(DiD1)
```

\subsection{(iv)} \label{q2_4}

```{r}
DiD2= feols(lnm_rate ~ indicator*treated| year + treated, data = dfData, se="standard")
msummary(list(DiD1,DiD2), stars = c('*' = .1, '**' = .05, '***' = .01))
```

(For (iii) and (iv), I took out the group-specific and time-specific effect. However, if you want to interpret those you can take out the "\|year + treated" part in the model specification)

\subsection{(v)} \label{q2_5}

For this question, we can take out the two-way fixed effect of group and year and create an interaction variable of year and treated group, making the year 1936 the reference year and thus normalize its coefficients to 0. This way, we estimate the effect of the treatment before and after the event with considering the different periods around the particular year.

```{r}
es <- feols(lnm_rate ~ i(year, treated, ref = 1936)|year+treated, data = dfData)
summary(es)
coefplot(es, ylab = "Coefficient", innerCI=5, outerCI=7, lwdInner=5, lwdOuter=7)
```

From the plot we can see that across the groups (states), before the 

\subsection{(vi)} \label{q2_6}

We need to cluster standard errors at he fixed effect levels: namely year and treated

\subsection{(vii)} \label{q2_7}

For this, we can use the Wald test to check on the pre-trends from the event study model before the actual treatment period.

```{r}
# https://lrberge.github.io/fixest/reference/wald.html
wald(es, keep=c("year::1925:treated","year::1926:treated","year::1927:treated","year::1928:treated","year::1929:treated","year::1930:treated","year::1931:treated","year::1932:treated","year::1933:treated","year::1934:treated","year::1935:treated"))
```

We reject the null hypothesis, which means that there is enough evidence which suggests that the parallel trends condition is not satisfied. This means that the decreasing trend in scarlet fever was stronger anyway compared to the tuberculosis and regardless of the treatment (the appearance of a drug effective for the former but not for the latter). Then we can use the placebo test, in which we pick fake treatment periods before the actual treatment period and see if there is a significant effect. In theory, if the common trends assumption was satisfied, the placebo (fake) treatment should not be significant, only the real treatment effect should yield a significant effect.

```{r}
# Create fake indicator variables
dfData$D_fake1 <- ifelse(dfData$year >=1928, 1, 0)
dfData$D_fake2 <- ifelse(dfData$year >=1930, 1, 0)
dfData$D_fake3 <- ifelse(dfData$year >=1932, 1, 0)
dfData$D_fake4 <- ifelse(dfData$year >=1934, 1, 0)

# Test fake models
DiD1_fake = feols(lnm_rate ~ D_fake1*treated|year + treated, data = dfData, cluster = "treated^year")
DiD2_fake = feols(lnm_rate ~ D_fake2*treated|year + treated, data = dfData, cluster = "treated^year")
DiD3_fake = feols(lnm_rate ~ D_fake3*treated|year + treated, data = dfData, cluster = "treated^year")
DiD4_fake = feols(lnm_rate ~ D_fake4*treated|year + treated, data = dfData, cluster = "treated^year")
msummary(list(DiD1_fake,DiD2_fake,DiD3_fake,DiD4_fake), stars = c('*' = .1, '**' = .05, '***' = .01))
```

Using the placebo test, we also find that the parallel trend pre-treatment period is not satisfied, because the fake treatment periods yield significant treatment effects as well. Therefore using event studies alone is not warranted and one should consider differences-in-differences or differences-in-differences-in-differences instead.
